---
title: "EDA"
output: github_document
---


```{r}
library(tidyverse)
```

## load the data

```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```

## group by 
```{r}
weather_df%>%
  group_by(name,month)
```

## counting things

count month/name observations

```{r}
weather_df%>%
  group_by(name, month) %>%
  summarize(n_obs = n())
```

we can use count
```{r}
weather_df%>%
  count(name, month, name = "n_obs")
```

## never use base R's table!!!

```{r, eval=FALSE}
weather_df%>%
  pull(month)%>%
  table()
```

other helpful counters

```{r}
weather_df%>%
  group_by(month)%>%
  summarize(
    n_obs = n(),
    n_days = n_distinct(date)
  )
```

## a diagression on 2x2 tables

```{r}
weather_df%>%
  filter(name != "Waikiki_HA")%>%
  mutate(
    cold = case_when(
      tmax< 5~"cold",
      tmax>=5~"not_cold",
      TRUE ~""
    )
  )%>%
  group_by(name,cold)%>%
  summarize(count = n())
```

```{r}
weather_df%>%
  filter(name != "Waikiki_HA")%>%
  mutate(
    cold = case_when(
      tmax< 5~"cold",
      tmax>=5~"not_cold",
      TRUE ~""
    )
  )%>%
  janitor::tabyl(name,cold)
```

## general summarises

YOU CAN DO LOTS OF summarises
```{r}
weather_df%>%
  group_by(name, month)%>%
  summarize(
    mean_tmax = mean (tmax, na.rm = TRUE),
    mean_prcp = mean (prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  )
```

This is a dataframe!

```{r}
weather_df%>%
  group_by(name, month)%>%
  summarize(
    mean_tmax = mean (tmax, na.rm = TRUE),
    mean_prcp = mean (prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
  )%>%
  ggplot(aes(x = month, y = mean_tmax, color = name))+
  geom_point()+
  geom_line()
  
```

suppose we want to summarize many columns.

```{r}
weather_df%>%
  group_by(name, month)%>%
  summarize(across(prcp:tmin,mean))
```

Reminder: sometimes your results are easier to read in another format.


```{r}
weather_df%>%
  group_by(name, month)%>%
  summarize(mean_tmax =mean(tmax))%>%
  pivot_wider(
    names_from = name, 
    values_from = mean_tmax
  )%>%
  knitr::kable(digits = 1)
```


##group_by and mutate

```{r}
weather_df%>%
  group_by(name)%>%
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax-mean_tmax
  )%>%
  ggplot(aes(x = date, y = centered_tmax, color= name))+
  geom_point()
```

what about window functions

ranking

```{r}
weather_df%>%
  group_by(name,month)%>%
  mutate(temp_rank = min_rank(tmax))
```

lag

```{r}
weather_df%>%
  group_by(name)%>%
  mutate(temp_change =tmax- lag(tmax))%>%
  summarize(
    temp_change_max = max(temp_change, na.rm = TRUE),
    temp_change_sd = sd(temp_change,na.rm = TRUE)
  )
```

## Quick note

summarize only gets you so far



